### 2012-03-25 paper_draft.tex
Fixed colons and lowercases. Put footnotes into references according to http://blogs.plos.org/mfenner/2011/07/21/how-to-formally-cite-a-blog-post/

If the genotypes for more then one individual are known, the analysis of
allele frequencies can be used to associate SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
(GWAS) make use of statistics to compare the alleles of patients to the alleles of healthy controls. This
enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
get an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
alone had over 100,000 customers\footnote{http://spittoon.23andme.com/2011/06/15/23andme-2011-state-of-the-database-address/} 
- the company recognizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
traits and diseases. With the consent of the customer the data is used for association studies. 23andMe has published several 
articles in which it replicates known findings, but also finds new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}. 
Over 30,000 23andme-customers have participated in those association studies.

Although companies like 23andMe are willing to use the data for scientific research,
it is not easy for individual scientists to access the data. 
The generation of biomedical data by private companies raises concerns about privacy\footnote{https://www.23andme.com/about/privacy/}, 
liabilty and consent \cite{Caulfield2011}. 
Nevertheless, individual customers are willingly sharing their data. Most do so by uploading their data to
their personal website or to open software repositories like \textit{GitHub}. 
This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all 
the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is 
most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
like the \textit{Personal Genome Project}, still do not allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.

@@

Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations 
for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they 
fear discrimination (mean difference = 1.52, SE = 0.322), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324), 
because they fear consequences for their family members (mean difference = 1.146, SE = 0.32) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357). In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist: Those who would share their data with their DTC company as long as it did not share the information agree less strongly than those who would not share the data with both fear of discrimination as a motive for not sharing (mean difference = -0.615, SE = 0.345) as well as fear of a breach of privacy (mean difference = -0.668, SE = 0.346). Those who would share their data under any circumstances are even less likely to agree with these motives than those who would share only if their DTC company did not share the information (fear of discrimination: mean value = -0.906, SE = 0.182; breach of privacy: mean difference = -1.203, SE = 0.183).

@@

The mean number of users that have entered their variations for a single phenotype 
is 14.36 (SD 12.65), the median is 10. The distribution of how many users have 
entered their data per phenotype can be seen in figure \ref{Figure2_label}. The phenotype provided 
by the most users is the eye color, which has been entered by 54 users. There are 
two phenotypes which have so far only been provided by a single user: 
The SAT Writing score and triglyceride-levels.

@@

There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one 
could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
variations that have not been available from the start. Both methods have a disadvantage: In either case it makes it harder 
for users to enter their data and thereby raises the bar for participation, which ultimately could lead to less data entered. 
Facing this trade-off, we decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS 
with the data to perform more quality control.

>>>>

If the genotypes for more then one individual are known, the analysis of
allele frequencies can be used to associate SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
(GWAS) make use of statistics to compare the alleles of patients to the alleles of healthy controls. This
enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising. Over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
get an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
alone had over 100,000 customers \cite{23andMe2011}
- the company recognizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
traits and diseases. With the consent of the customer the data is used for association studies. 23andMe has published several 
articles in which it replicates known findings, but also finds new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}. 
Over 30,000 23andme-customers have participated in those association studies.

Although companies like 23andMe are willing to use the data for scientific research,
it is not easy for individual scientists to access the data. 
The generation of biomedical data by private companies raises concerns about privacy \cite{23andMe2012}, 
liabilty and consent \cite{Caulfield2011}. 
Nevertheless, individual customers are willingly sharing their data. Most do so by uploading their data to
their personal website or to open software repositories like \textit{GitHub}. 
This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all 
the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is 
most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
like the \textit{Personal Genome Project}, still do not allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.

@@

Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations 
for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they 
fear discrimination (mean difference = 1.52, SE = 0.322), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324), 
because they fear consequences for their family members (mean difference = 1.146, SE = 0.32) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357). In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist. Those who would share their data with their DTC company as long as it did not share the information agree less strongly than those who would not share the data with both fear of discrimination as a motive for not sharing (mean difference = -0.615, SE = 0.345) as well as fear of a breach of privacy (mean difference = -0.668, SE = 0.346). Those who would share their data under any circumstances are even less likely to agree with these motives than those who would share only if their DTC company did not share the information (fear of discrimination: mean value = -0.906, SE = 0.182; breach of privacy: mean difference = -1.203, SE = 0.183).

@@

The mean number of users that have entered their variations for a single phenotype 
is 14.36 (SD 12.65), the median is 10. The distribution of how many users have 
entered their data per phenotype can be seen in figure \ref{Figure2_label}. The phenotype provided 
by the most users is the eye color, which has been entered by 54 users. There are 
two phenotypes which have so far only been provided by a single user: 
the SAT Writing score and triglyceride-levels.

@@

There are two possible solutions to this problem: one could only allow some (trusted) users to enter new phenotypes or one 
could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
variations that have not been available from the start. Both methods have a disadvantage: In either case it makes it harder 
for users to enter their data and thereby raises the bar for participation, which ultimately could lead to less data entered. 
Facing this trade-off, we decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS 
with the data to perform more quality control.
